Modeling trend in temperature volatility using generalized LASSO

### 2018-05-13 nips-2018/paper.tex
reorganization and adding unique bib

\usepackage[showonlyrefs]{mathtools}

@@

@@

@@

In this project, we consider the problem of detecting the trend in the temperature volatility. All analyses are performed on a sub-set of the European Centre for Medium-Range Weather Forecasts (ECMWF) ERA-40 dataset \citep{uppala_era-40_2005}. This dataset include the temperature measurements over a grid over the earth from 1957 to 2002.

@@

\section{Data Exploration}
This section is devoted to exploring some of the properties of the ERA-40 surface temperature data set. The goal here is to demonstrate some of the difficulties in modeling the trend in the temperature volatility and motivate our methodology for doing so.

\autoref{fig:cities_ts} shows the time-series of the temperature of three cities: Bloomington (USA), San Diego (USA) and Manaus (Brazil). The time-series of Bloomington and San Diego show clear cyclic behavior. However, while it seems (qualitatively) that these cycles can be modeled by a sinusoidal function for Bloomington, the same is not true for San Diego. Also, the amplitude of the cycles changes from some years to others. The time-series of Manaus does not show any regular cyclic behavior. This demonstrates the first difficulty in analyzing the variance of this data: to analyze the variance, we first need to remove the cyclic terms from all time-series. However, there is a lot of variations in the cyclic behavior of the time-series of different locations. In addition, some of these cycles cannot be easily modeled by a parametric function \footnote{One might try to model the cycles by the summation of sinusoidal terms with different frequencies. However, for some time-series this may need many terms to be included in the summation to achieve a reasonable level of accuracy. In addition, this model cannot capture the non-stationarity in the cycles.}. To overcome these issues, we will use a non-parametric approach to remove the cyclic terms from the time-series and de-trend them. This approach, called \textit{$\ell_1$-trend filtering} is explained in the next section.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/cities_ts}}
		\caption{Time-series of the temperature (in Kelvin) of three cities.}
		\label{fig:cities_ts}
	\end{center}
	\vskip -0.2in
\end{figure}

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/bloom_trend_resid}}
		\caption{Left: Time-series of the temperature of Bloomington (blue) and the estimated trend and cycles obtained from the $\ell_1$-trend filtering (red). Right: the same time-series after removing the cyclic terms and de-trending using $\ell_1$-trend filtering.}
		\label{fig:bloom_detrended}
	\end{center}
	\vskip -0.2in
\end{figure}

\section{$\ell_1$-trend filtering for de-trending}

$\ell_1$-trend filtering was proposed by \citep{kim_$ell_1$_2009} as a solution to the following problem: a time-series $y_t$, $t=1,...,T$ is observed and we believe that it consists of a smooth trend and a stochastic component and the goal is to estimate the trend. The estimated trend is desired to be smooth, but at the same time, we seek small estimated residuals (stochastic component). These two objectives cannot be achieved simultaneously and so a trade-off between them should be made. In $\ell_1$-trend filtering, this trade-off is formulated as the following convex optimization problem:

\begin{equation}
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda \sum_{t=1}^{T-2} |\beta_t-2\beta_{t+1}+\beta_{t+2}|
\end{equation}

or equivalently:

\begin{equation}
\min_{\beta} \frac{1}{2} \lVert y-\beta \lVert_2^2+\lambda \lVert Dh \lVert_1
\label{eq:l1tf}
\end{equation}

 where $||u||_1$ is the $\ell_1$ norm of the vector $u$. Here, $\beta$ is the vector of the free parameters of the model and there is one parameter per observation. The \textit{penalty matrix} is:

\begin{equation}
D=\begin{bmatrix}
1 & -2 & 1 &  &  &  &\\ 
& 1 & -2 & 1 &  &  &\\ 
&  & \ddots & \ddots & \ddots  &  &\\ 
&  & & 1 & -2 & 1 &  \\ 
&  &  &   & 1 & -2 & 1 
\end{bmatrix}
\label{eq:d_matrix}
\end{equation}

The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$. Specifically, with the penalty matrix \eqref{eq:d_matrix}, the estimated $\beta$ will be piecewise linear \citep{kim_$ell_1$_2009}. The trade-off between the two aforementioned objectives is determined by the constant $\lambda$. \citep{kim_$ell_1$_2009} proposed a specialized primal-dual interior point (PDIP) method for solving the optimization problem \eqref{eq:l1tf} with the penalty matrix \eqref{eq:d_matrix}.

The results of applying $\ell_1$-trend filtering on the Bloomington time-series are shown in \autoref{fig:bloom_detrended}. The left panel shows the original time-series (blue) together with the estimated $\beta$ (red) and the right panel shows the residuals $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are piecewise linear. The number of linear segments is determined by the value of $\lambda$. Here, we chose $\lambda=500$. The choice of $\lambda$ is a model selection problem and we will return to it in the next section ???.

\section{$\ell_1$-trend filtering for estimating variance of a time-series}
\label{sec:l1tf_var}

%TODO: talk about the motivation for using non-parametric based on the figure of the residuals.

In this section, inspired by the $\ell_1$-trend filtering algorithm, we propose a non-parametric model for estimating the variance of a time-series. To this end, we assume that at each time step $t$, there is a hidden variable $h_t$ such that conditioned on $h_t$ the observations $y_t$ are independent normal variables with zero mean and variance $\exp(h_t)$. The conditional log-likelihood of the observed data in this model is $l(y|h)=c-\sum_{t=1}^Th_t-y_t^2e^{-h_t}$, where $c$ is a constant that does not depend on $y$ and $h$. Crucially, we assume that the hidden variables $h_t$ vary smoothly. To impose this assumption, we estimate $h_t$ by solving the penalized conditional log-likelihood:

\begin{equation}
\min_h -l(y|h)+\lambda \lVert Dh \lVert_1
\label{eq:l1tf_var}
\end{equation}

 where $D$ is the same as in \eqref{eq:d_matrix}.

%TODO: explain the objective more. give the AR(1) example. Explain what you loses by this assumption (ACF,forecasting). Also explain that the covariace matrix is diagonal so it cannot capture the covariance structure. But incontrast to spatial stat literature, it does not make any assumption on estimated variances. Compare to Hallac et al and Lingren et al.

We solve this optimization problem using the PDIP method \footnote{We use PDIP implemented in the cvxopt Python package \citep{andersen_cvxopt:_2013}.}. First, we note that this is a generalized LASSO problem \citep{tibshirani_solution_2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \lVert Dx \lVert_1$ is:

\begin{equation}
\begin{aligned}
\min_\nu f^*(-D^T\mathsf{\nu}) \\
s.t. \,\,\, \lVert \mathsf{\nu} \lVert_\infty \le \lambda
\end{aligned}
\end{equation}

 where $D^t$ is the transpose of $D$, and $f^*(\cdot)$ is the \textit{conjugate} of $f$ defined by: $f^*(u)=\max_x u^tx-f(x)$. It is simple to show that the conjugate of $f(h)=-l(y|h)$ is:

\begin{equation}
f^*(\mathsf{u})=\sum_t (u_t-1)\log\frac{y_t^2}{1-u_t} + u_t-1
\label{eq:conj}
\end{equation}

The red curve in the left panel of \autoref{fig:bloom_estimatedSD} shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of the time-series of Bloomington. To reduce the number of time-steps in this figure and in the remainder of the paper we work on the weekly averaged of the data.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}}
		\caption{Left: The residuals of the time-series of Bloomington (averaged weekly) and the estimated SD obtained from the method of \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and their annual average (black) without imposing the long horizon penalty. Right: the same as middle panel but here the long horizon penalty is imposed. See the text for more details.}
		\label{fig:bloom_estimatedSD}
	\end{center}
	\vskip -0.2in
\end{figure}

\section{Extension to spatio-temporal data}
\label{sec:exten}

@@

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=.5\columnwidth]{Figures/data_cube}}
		\caption{The cube represents the global variable $z$ in space and time. The sub-cubes specified by the white lines are $x_i$.}
		\label{fig:data_cube}
	\end{center}
	\vskip -0.2in
\end{figure}

@@

\section{Results}
As it was mentioned before, the algorithm proposed in \autoref{sec:consOpt} is appropriate only if we parallelize it over multiple machines and so we do not pursue it further here. All the results reported in this section are obtained using the linearize ADMM algorithm of \autoref{sec:linADMM}. We applied this algorithm on a subset of the ERA-40 dataset. The data is the 2 meter temperature measured daily at 12 p.m from August 31 of 1992 to 2002. To reduce the noise we first computed the weekly average of this data. To further reduce the size of the data, we will only analyze the data of the locations inside a rectangle extended from (\ang{58},\ang{226}) to (\ang{22},\ang{302}). This rectangle covers the united states. All the computations were performed on a machine with four 3.20GHz Intel i5-3470 cores.

\subsection{Convergence}

\autoref{fig:loss_vs_iter} shows the value of the loss function \eqref{eq:l1tf_var_st_mat} as a function of iteration for $\lambda_t=1$ and $\lambda_s=0$. We used the following rule to determine when to stop the optimization: the optimization was stopped if the value of the loss did not improve by at least 0.1\% in 1000 trials. As we can see, the algorithm converged in about 2000 iterations. This took about 11 minutes. Our experiments showed that the convergence speed depends on the value of $\lambda_t$ and $\lambda_s$. Also, if we use the solution obtained for smaller values of these parameters as the initial value for the larger values (\textit{warm start}), the converges speed improves.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=1\columnwidth]{Figures/loss_vs_iter}}
		\caption{Loss as a function of iteration for $\lambda_t=1$ and $\lambda_s=0$.}
		\label{fig:loss_vs_iter}
	\end{center}
	\vskip -0.2in
\end{figure}

\subsection{Model selection}
One common method for choosing the penalty parameters in the Lasso problems is to find the solution for a range of the values of these parameters and then choose the values which minimize a model selection criterion. However, such analyses needs the computation of the degrees of freedom (df). Several previous work have investigated the df in generalized lasso problems \citep{tibshirani_degrees_2012,hu_dual_2015,zeng_geometry_2017}. However, all these studies have considered the linear regression problem and, to the best of our knowledge, the problem of computing the df for generalized lasso with general objective function has not been considered yet.

@@

We computed the optimal solution for all the combinations of the following sets of values: $\lambda_t \in \{1,5,10,20\} \, \, , \lambda_s \in \{0,.1,1,5,10\}$. The value of $\mathscr{L}(\lambda_t,\lambda_s)$ is shown in \autoref{fig:modelSelect}. This function is minimized at $\lambda_t=5$ and $\lambda_s=1$. All the analyses in the next section are performed on the solution for these values.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=1\columnwidth]{Figures/modelSelect}}
		\caption{Value of $\mathscr{L}(\lambda_t,\lambda_s)$ for several values of $\lambda_t$ and $\lambda_s$. }
		\label{fig:modelSelect}
	\end{center}
	\vskip -0.2in
\end{figure}

\subsection{Analysis of trend of temperature volatility}
\autoref{fig:estimatedSD} shows the detrended data, the estimated standard deviation and the yearly average of these estimates for two cities in the united states: Bloomington (left) and San Diego (right). The estimated SD captures the periodic behavior in the variance of the time-series. In addition, the number of linear segments changes adaptively in each time window depending on how fast the variance is changing.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=1\columnwidth]{Figures/estimatedSD}}
		\caption{Detrended data and the estimated SD for Bloomington (left) and San Diego (right).}
		\label{fig:estimatedSD}
	\end{center}
	\vskip -0.2in
\end{figure}

The yearly average of the estimated SD captures the trend in the temperature volatility. For example, we can see that in Bloomington, there is a small positive trend. To determine how the volatility has changed in each location, we subtract the average of the estimated variance in 1992 from the average in the following years and compute their sum. The value of this change in the variance in each location is depicted in the right panel of \autoref{fig:avg_change_estimatedSD}. The left panel of this figure, shows the average estimated variance in each location.

\begin{figure}[ht]
	\begin{subfigure}
	  \centering
	  \includegraphics[width=.5\linewidth]{Figures/avg_estimatedVar}
	\end{subfigure}%
	\begin{subfigure}
	  \centering
	  \includegraphics[width=.5\linewidth]{Figures/avg_change_estimatedVar}
	\end{subfigure}%
	\caption{Left: the average of the estimated variance in all locations in the grid over the united states. Right: the change in the variance from 1992 to 2002.}
	\label{fig:avg_change_estimatedSD}
\end{figure}

It is interesting to note that the trend in volatility is almost zero over the oceans. The most positive trend can be observed in the south-east and the most negative trend has happened in the north-east.

Finally, \autoref{fig:hist_avg_change} shows the histogram of the changes shown in the right panel of \autoref{fig:avg_change_estimatedSD}. As we can see, there are two peaks in this histogram: one for locations with no trend (which includes the points on the oceans), and one peak for locations with positive trend. The locations with negative trend form the long tail of the histogram.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=1\columnwidth]{Figures/hist_avg_change}}
		\caption{Histogram of the change in the estimated variances.}
		\label{fig:hist_avg_change}
	\end{center}
	\vskip -0.2in
\end{figure}

@@

\bibliographystyle{abbrvnat}
\small
\bibliography{references}

>>>>

\usepackage{mathtools}
\mathtoolsset{showonlyrefs}

@@

\usepackage{xcolor}
\newcommand{\alert}[1]{\textcolor{red}{TODO: #1}}

@@

Nonparametric variance estimation for spatio-temporal data.

\subsection{Motivating applications}

@@

In this project, we consider the problem of detecting the trend in the temperature volatility. All analyses are performed on a sub-set of the European Centre for Medium-Range Weather Forecasts (ECMWF) ERA-40 dataset \citep{uppala_era-40_2005}. This dataset include the temperature measurements over a grid over the earth from 1957 to 2002.~\citep{VasseurDeLong2014,TrenberthZhang2014,StatenKahn2016,Screen2014,FischerBeyerle2013}

@@

Also neuroscience.

\subsection{Related work}

Mention ~\citep{KimKoh2009,HallacPark2017}. Also, ~\citep{Tibshirani2014,TibshiraniTaylor2011}. Find Sharpnack Tibshirani. ARCH/GARCH.

\citep{SadhanalaWang2017,LinSharpnack2017,WangSharpnack2016}~\citep{RamdasTibshirani2016}

\subsection{Main contributions}

\section{$\ell_1$-trend filtering for estimating variance of a time-series}

$\ell_1$-trend filtering was proposed by \citep{kim_$ell_1$_2009} as a method for estimating a smooth, time-varying trend. It is formulated as the optimization problem
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that it consists of a smooth trend and a stochastic component and the goal is to estimate the trend. The estimated trend is desired to be smooth, but at the same time, we seek small estimated residuals (stochastic component). These two objectives cannot be achieved simultaneously and so a trade-off between them should be made. In $\ell_1$-trend filtering, this trade-off is formulated as the following convex optimization problem:
\begin{equation}
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda \sum_{t=1}^{T-2} |\beta_t-2\beta_{t+1}+\beta_{t+2}|
\end{equation}
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \lVert y-\beta \lVert_2^2+\lambda \lVert D\beta\lVert_1
\label{eq:l1tf}
\end{equation}
 where $y_t$ is an observed time-series, $\beta$ is the smooth trend, $D$ is a $T-2\times T$ matrix, and $\lambda$ is a tuning parameter which balances fidelity to the data (small errors in the first term) with a desire for smoothness. 
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise linear. \citep{kim_$ell_1$_2009} proposed a specialized primal-dual interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a statistical perspective, \eqref{eq:l1tf} is a constrained maximum likelihood problem with independent observations from a normal distribution with common variance, $y_t \sim \mbox{N}(\beta_t, \sigma^2)$, subject to a piecewise linear constraint on $\beta$.

% The results of applying $\ell_1$-trend filtering on the Bloomington time-series are shown in \autoref{fig:bloom_detrended}. The left panel shows the original time-series (blue) together with the estimated $\beta$ (red) and the right panel shows the residuals $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are piecewise linear. The number of linear segments is determined by the value of $\lambda$. Here, we chose $\lambda=500$. The choice of $\lambda$ is a model selection problem and we will return to it in the next section ???.

\subsection{Extension to variance estimation}
\label{sec:l1tf_var}

Inspired by the $\ell_1$-trend filtering algorithm, we propose a non-parametric model for estimating the variance of a time-series. To this end, we assume that at each time step $t$, there is a hidden variable $h_t$ such that conditioned on $h_t$ the observations $y_t$ are independent normal variables with zero mean and variance $\exp(h_t)$. The negative log-likelihood of the observed data in this model is $l(y|h) \propto -\sum_{t=1}^Th_t - y_t^2e^{-h_t}$. Crucially, we assume that the hidden variables $h_t$ vary smoothly. To impose this assumption, we estimate $h_t$ by solving the penalized, negative log-likelihood:
\begin{equation}
\min_h -l(y|h)+\lambda \lVert Dh \lVert_1
\label{eq:l1tf_var}
\end{equation}
 where $D$ has the same structure as above.

\alert{explain the objective more. give the AR(1) example. Explain what you loses by this assumption (ACF,forecasting). Also explain that the covariace matrix is diagonal so it cannot capture the covariance structure. But incontrast to spatial stat literature, it does not make any assumption on estimated variances. Compare to Hallac et al and Lingren et al.}

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the PDIP algorithm (as in, e.g., \texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). First, we note that this is a generalized LASSO problem \citep{tibshirani_solution_2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \lVert Dx \lVert_1$ is: 
\begin{align}
\min_\nu&\quad f^*(-D^\top\nu) \\
\mbox{s.t.}&\quad \lVert \nu \lVert_\infty \le \lambda
\end{align}
 where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x u^tx-f(x)$. It is simple to show that 
\begin{equation}
f^*(\mathsf{u})=\sum_t (u_t-1)\log\frac{y_t^2}{1-u_t} + u_t-1.
\label{eq:conj}
\end{equation}

\subsection{Extension to spatio-temporal data}
\label{sec:exten}

@@

\begin{figure}[tb]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=.5\columnwidth]{Figures/data_cube}}
		\caption{The cube represents the global variable $z$ in space and time. The sub-cubes specified by the white lines are $x_i$.}
		\label{fig:data_cube}
	\end{center}
	\vskip -0.2in
\end{figure}

@@

\section{Empirical evaluation}

\subsection{Simulations}
\label{sec:simulations}

To analyze the proposed optimization methods, we first fit the model on some synthetic spatio-temporal data. The observations at all time steps and all locations were generated from independent Gaussian random variables with zero mean. However, the variance of these random variables changes smoothly in time and space. Specifically, the observation at time step $t$ and at location $(r,c)$ on the grid is generated from a Gaussian $N(0,\sigma^2(t,r,c))$, where:

\begin{equation}
\begin{aligned}
\sigma(t,r,c) & =\sum_{s=1}^{S} W_s(t) \cdot \exp\bigg( \frac{(r-r_s)^2+(c-c_s)^2}{2\sigma_s^2} \bigg) \\
W_s(t) & =\alpha_s \cdot t + \exp(\sin(2\pi\omega_s t+\phi_s)) 
\label{eq:sourceVar}
\end{aligned}
\end{equation}

In words, the variance at each time and location is computed as the weighted sum of $S$ bell-shaped functions where the weights are time-varying functions. Specifically, the weights consist of a linear trend $\alpha_s \cdot t$ and a periodic term $\beta_s \cdot \sin(2\pi\omega_s t+\phi_s)$. The bell-shaped functions impose the spatial smoothness, and the linear trend and the periodic terms enforce the temporal smoothness. We simulated the data on a 5 by 7 grid and for 780 time steps. We also used $S=4$. The parameters of the variance function are shown in  \autoref{tab:sim_params}. The value of the variance function for all locations on the grid and at $t=25$ and $t=45$ is shown in \autoref{fig:true_var_spatial}. Also, the variance for all time steps and at the location (0,0) on the grid is shown in \autoref{fig:true_fitted_var}. The linear trend and the period term can be seen clearly in this figure.

\begin{table}[tb]
	\caption{Parameters used to simulate data}
	\label{tab:sim_params}
	\vskip 0.15in
	\begin{center}
		\begin{small}
			\begin{sc}
				\begin{tabular}{ccccccc}
					\hline

					$s$ & $r_s$ & $c_s$ & $\sigma_s$ &$\alpha_s$ & $\omega_s$ & $\phi_s$\\
					\hline
					1 & 0 & 0 & 5 & 0.5 & 0.121 & 0 \\
					2 & 0 & 5 & 5 & 0.1 & 0.121 & 0 \\
					3 & 3 & 0 & 5 & -0.5 & 0.121 & $\pi/2$ \\
					4 & 3 & 5 & 5 & -0.1 & 0.121 & $\pi/2$ \\
					\hline
				\end{tabular}
			\end{sc}
		\end{small}
	\end{center}
	\vskip -0.1in
\end{table}

\begin{figure}[tb]
	\centering	
	\includegraphics[width=.7\linewidth]{Figures/true_var_spatial}

	\caption{Variance function at $t=25$ (left) and $t=45$ (right)} \label{fig:true_var_spatial}
\end{figure}

The left panel of \autoref{fig:true_fitted_var} shows the convergence of the two algorithms. Each iteration of the linearized algorithm takes about 0.01 seconds on average while each iteration of the consensus ADMM takes about 20 seconds.

We estimated the linearized ADMM for all combinations of values of $\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in \{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For each pair, we then compute the mean absolute error (MAE) between the estimated variance and the true variance at all locations and all time steps. For $\lambda_t=5$ and $\lambda_s=0.1$ MAE was minimized. The right panel of \autoref{fig:true_fitted_var} shows the true and the estimated standard deviation at location (0,0) using $\lambda_s=0.1$ and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green). As we can see, larger than optimal value of $\lambda_t$ leads to estimated values which are ``too smooth".

\begin{figure}[!h]
	\centering	
	\includegraphics[width=.4\linewidth]{Figures/convergence}			
	\includegraphics[width=.4\linewidth]{Figures/true_fitted_var}		
	\caption{Left: Convergence speed of linearized and consensus ADMM. Right: The true (orange) and estimated standard deviation function at the location (0,0). The estimated values are obtained using linearized ADMM with $\lambda_s=0.1$ and two values of $\lambda_t$: $\lambda_t=5$ (blue) and $\lambda_t=100$ (green).} \label{fig:true_fitted_var}
\end{figure}

\subsection{Data analysis}
As it was mentioned before, the algorithm proposed in \autoref{sec:consOpt} is appropriate only if we parallelize it over multiple machines and so we do not pursue it further here. All the results reported in this section are obtained using the linearize ADMM algorithm of \autoref{sec:linADMM}. We applied this algorithm on a subset of the ERA-40 dataset. The data is the 2 meter temperature measured daily at 12 p.m from August 31 of 1992 to 2002. To reduce the noise we first computed the weekly average of this data. To further reduce the size of the data, we will only analyze the data of the locations inside a rectangle extended from (\ang{58},\ang{226}) to (\ang{22},\ang{302}). This rectangle covers the united states. All the computations were performed on a machine with four 3.20GHz Intel i5-3470 cores.

\paragraph{Data Exploration}
The red curve in the left panel of \autoref{fig:bloom_estimatedSD} shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of the time-series of Bloomington. To reduce the number of time-steps in this figure and in the remainder of the paper we work on the weekly averaged of the data.

\begin{figure}[tb]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}}
		\caption{Left: The residuals of the time-series of Bloomington (averaged weekly) and the estimated SD obtained from the method of \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and their annual average (black) without imposing the long horizon penalty. Right: the same as middle panel but here the long horizon penalty is imposed. See the text for more details.}
		\label{fig:bloom_estimatedSD}
	\end{center}
	\vskip -0.2in
\end{figure}

This section is devoted to exploring some of the properties of the ERA-40 surface temperature data set. The goal here is to demonstrate some of the difficulties in modeling the trend in the temperature volatility and motivate our methodology for doing so.

% \autoref{fig:cities_ts} shows the time-series of the temperature of three cities: Bloomington (USA), San Diego (USA) and Manaus (Brazil). The time-series of Bloomington and San Diego show clear cyclic behavior. However, while it seems (qualitatively) that these cycles can be modeled by a sinusoidal function for Bloomington, the same is not true for San Diego. Also, the amplitude of the cycles changes from some years to others. The time-series of Manaus does not show any regular cyclic behavior. This demonstrates the first difficulty in analyzing the variance of this data: to analyze the variance, we first need to remove the cyclic terms from all time-series. However, there is a lot of variations in the cyclic behavior of the time-series of different locations. In addition, some of these cycles cannot be easily modeled by a parametric function \footnote{One might try to model the cycles by the summation of sinusoidal terms with different frequencies. However, for some time-series this may need many terms to be included in the summation to achieve a reasonable level of accuracy. In addition, this model cannot capture the non-stationarity in the cycles.}. To overcome these issues, we will use a non-parametric approach to remove the cyclic terms from the time-series and de-trend them. This approach, called \textit{$\ell_1$-trend filtering} is explained in the next section.

% \begin{figure}[tb]
% 	\vskip 0.2in
% 	\begin{center}
% 		\centerline{\includegraphics[width=\columnwidth]{Figures/cities_ts}}
% 		\caption{Time-series of the temperature (in Kelvin) of three cities.}
% 		\label{fig:cities_ts}
% 	\end{center}
% 	\vskip -0.2in
% \end{figure}

\begin{figure}[tb]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/bloom_trend_resid}}
		\caption{Left: Time-series of the temperature of Bloomington (blue) and the estimated trend and cycles obtained from the $\ell_1$-trend filtering (red). Right: the same time-series after removing the cyclic terms and de-trending using $\ell_1$-trend filtering.}
		\label{fig:bloom_detrended}
	\end{center}
	\vskip -0.2in
\end{figure}

\paragraph{Convergence}

We used the following rule to determine when to stop the optimization: the optimization was stopped if the value of the loss did not improve by at least 0.1\% in 1000 trials. As we can see, the algorithm converged in about 2000 iterations. This took about 11 minutes. Our experiments showed that the convergence speed depends on the value of $\lambda_t$ and $\lambda_s$. Also, if we use the solution obtained for smaller values of these parameters as the initial value for the larger values (\textit{warm start}), the converges speed improves.

\paragraph{Model selection}
One common method for choosing the penalty parameters in the Lasso problems is to find the solution for a range of the values of these parameters and then choose the values which minimize a model selection criterion. However, such analyses needs the computation of the degrees of freedom (df). Several previous work have investigated the df in generalized lasso problems \citep{tibshirani_degrees_2012,hu_dual_2015,zeng_geometry_2017}. However, all these studies have considered the linear regression problem and, to the best of our knowledge, the problem of computing the df for generalized lasso with general objective function has not been considered yet.

@@

We computed the optimal solution for all the combinations of the following sets of values: $\lambda_t \in \{1,5,10,20\} \, \, , \lambda_s \in \{0,.1,1,5,10\}$. The best combination based on a held out set was $\lambda_t=5$ and $\lambda_s=1$. All the analyses in the next section are performed on the solution for these values.

\paragraph{Analysis of trend of temperature volatility}

The top row of \autoref{fig:avg_change_estimatedSD} shows the detrended data, the estimated standard deviation and the yearly average of these estimates for two cities in the US. The estimated SD captures the periodic behavior in the variance of the time-series. In addition, the number of linear segments changes adaptively in each time window depending on how fast the variance is changing. 
The yearly average of the estimated SD captures the trend in the temperature volatility. For example, we can see that in Bloomington, there is a small positive trend. To determine how the volatility has changed in each location, we subtract the average of the estimated variance in 1992 from the average in the following years and compute their sum. The value of this change in the variance in each location is depicted in the right panel of \autoref{fig:avg_change_estimatedSD}. The left panel of this figure, shows the average estimated variance in each location.

\begin{figure}[tb]
	  \centering
          \includegraphics[width=.9\columnwidth]{Figures/estimatedSD}\\
	  \includegraphics[width=.45\linewidth]{Figures/avg_estimatedVar}
	  \includegraphics[width=.45\linewidth]{Figures/avg_change_estimatedVar}
	\caption{Top row: Detrended data and the estimated SD for Bloomington (left) and San Diego (right). Bottom: the average of the estimated variance over the US (left) and the change in the variance from 1992 to 2002 (right)}
	\label{fig:avg_change_estimatedSD}
\end{figure} 
It is interesting to note that the trend in volatility is almost zero over the oceans. The most positive trend can be observed in the south-east and the most negative trend has happened in the north-east.

@@

\bibliographystyle{abbrvnat}
\small
\bibliography{nips-references}
